---
title: "2. Visualisation of DESeq2 results"
description: "PCA and clustered heatmap"
principal investigator: "Joaquín de Navascués"
researchers: "Aleix Puig-Barbé, Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: readable
    df_print: paged
    css: doc.css
---
```{r setup, echo=FALSE, cache=FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 7, fig.width = 8.5, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path='notebook_figs/', warning=FALSE, message=FALSE)
```


# 1 Preparation


**Libraries/utils:**
```{r libraries, warning=FALSE, message=FALSE}
if (!require("librarian")) install.packages("librarian")
librarian::shelf(
  # tidyverse
  tibble, stringr, tidyr, purrr, dplyr,
  # data other
  plyr, reshape2, santoku, calibrate, readxl,
  DESeq2, DescTools, matrixStats,
  # graphics
  ggplot2, ggthemes, ggtext, ggrepel,
  eulerr, RColorBrewer, cetcolor,
  # convenience
  here)

if(!exists("extract_regulated_sets", mode="function")) source("utils.R")
```

**Set working directory:**
```{r setwd}
if (Sys.getenv("RSTUDIO")==1) {
   # setwd to where the editor is, if the IDE is RStudio
  setwd( dirname(rstudioapi::getSourceEditorContext(id = NULL)$path) )
} else {
  # setwd to where the editor is in a general way - maybe less failsafe than the previous
  setwd(here::here())
  # the following checks that the latter went well, but assuming
  # that the user has not changed the name of the repo
  d <- str_split(getwd(),'/')[[1]][length(str_split(getwd(),'/')[[1]])]
  if (d != 'RNAseq-EmcDaSc-adult_midgut') { stop(
    paste0("Could not set working directory automatically to where this",
           " script resides.\nPlease do `setwd()` manually"))
    }
}
```

**To save images outside the repo (to reduce size):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(),'/')[[1]],-1),
                   paste0(tail(str_split(getwd(),'/')[[1]],1), '_figures')),
                 collapse='/')
dir.create(figdir, showWarnings = FALSE)
```


# 2 RNAseq results visualisation: Principal Components Analysis


## 2.1 Load and check expression data


We have seen how batch correction is necessary to show a meaningful PCA. Of the different expression measures, vst-normalised data is the most reasonable to make comparisons across samples.
```{r load-vst}
targets <- readRDS('output/targets.RDS')
vsd <- readRDS('output/vst_pseudocounts_batchCorrected.RDS')
```

When performing PCA on RNAseq data, it is often useful to filter genes with very low variance across experimental conditions. To test whether this makes sense in our case:
```{r visualise-variance, warning=FALSE}
# get per-gene variance and mean expression
var_vsd <- rowVars(assay(vsd), rm.na=TRUE)
avg_vsd <- rowMeans(assay(vsd))
df <- data.frame(Variance=var_vsd, Mean=avg_vsd)
# plot close to variance==0 with inset:
p1 <- ggplot(df, aes(x=Variance, y=Mean)) +
  geom_point(size=0.1, alpha=0.5, colour='#fe4b03') +
  geom_vline(xintercept=0, colour='black', size=0.1)
p2 <- ggplot(df, aes(x=Variance, y=Mean)) +
  geom_point(alpha=0.1) + xlim(0,0.1) + ylim(5,20) +
  theme(panel.background = element_rect(fill='grey90')) +
  geom_vline(xintercept=0, colour='#02ccfe')
p1 + annotation_custom(ggplotGrob(p2),
                       xmin=5, xmax=12,
                       ymin=5, ymax=25)
```

Some genes have very low variance (which is the whole point of the `vst` transformation), but none of them are zero.
There seems to be little point in drawing an arbitrary threshold, so I move on with the whole dataset.

## 2.2 PCA plot

```{r}
pca <- prcomp(t(assay(vsd)), center=TRUE)
scores <- data.frame(targets$sampleIDs, pca$x[,1:2])
xtitle <- paste0('**PC1** (', round(summary(pca)$importance['Proportion of Variance','PC1']*100,1), ' %)')
ytitle <- paste0('**PC2** (', round(summary(pca)$importance['Proportion of Variance','PC2']*100,1), ' %)')
ptitle <- 'Principal component scores' 
stitle <- '(*vst* pseudocounts, batch-corrected)'

ggplot(scores,
       aes(x = PC1, y = PC2,
           label=factor(targets$sampleIDs),
           colour=factor(targets$condition_md) )
       ) +
  # data
  geom_vline(xintercept=0, linetype='dashed', colour='grey60', linewidth=0.5) +
  geom_hline(yintercept=0, linetype='dashed', colour='grey60', linewidth=0.5) +
  geom_point(size=4) +
  geom_point(size=2, colour='white', alpha=0.5) + 
  geom_text_repel(size=4, alpha=0.75,
                  max.overlaps=Inf,
                  seed=42,
                  force=0.3, force_pull=1, 
                  box.padding=1, point.padding=0.5) +
  lims(x= c(-100, 150), y = c(-80, 80)) +
  # decorations
  theme_minimal(base_size=16) +
  labs(x=xtitle,
       y=ytitle,
       title=ptitle,
       subtitle=stitle) +
  scale_colour_discrete(name="condition") +
  theme(legend.text = element_markdown(size=10),
        axis.title.x = element_markdown(),
        axis.title.y = element_markdown(),
        plot.title = element_text(hjust=0.5),
        plot.subtitle = element_markdown(hjust=0.5, color="grey40"),
        axis.text.x = element_text(color="grey60"),
        axis.text.y = element_text(color="grey60"),
        axis.ticks.x = element_line(linewidth=0.5, linetype = "solid", color="grey60"),
        axis.ticks.y = element_line(linewidth=0.5, linetype = "solid", color="grey60"),
        axis.ticks.length=unit(-0.25, "cm"),
        panel.border = element_rect(linewidth=1, linetype = "solid", fill = NA),
        panel.grid = element_blank())

ggsave('PCA_vsd_batchcorr.pdf', plot = last_plot(), device = 'pdf',
       path = figdir, dpi = 300)
```


# 3 RNAseq results visualisation: Euler diagrams


# 3.1 Prepare data

```{r load-dge}
# experimental design and labels
targets <- readRDS('output/targets.RDS')
# DEG data
DaDaOE_deg <- readRDS('output/Control_vs_DaDaOE.RDS')
DaKD_deg <- readRDS('output/Control_vs_DaKD.RDS')
DaOE_deg <- readRDS('output/Control_vs_DaOE.RDS')
ScOE_deg <- readRDS('output/Control_vs_ScOE.RDS')
# gene symbols
dlist <- read.table(file="resources/gene_symbols.txt", header=TRUE)
names(dlist)[[1]] <- 'ensembl_gene_id'
rownames(dlist) <- dlist$ensembl_gene_id
```


Now, for each condition, we will need to get the lists of misregulated genes for a given threshold of fold change:
```{r extract_regulated_sets, warning = FALSE}
list_of_degs <- list(DaKD_deg, DaOE_deg, DaDaOE_deg, ScOE_deg)
names_degs <- list('daRNAi', 'da', 'da:da', 'scute')
fc_thresh <- 1.5
regs <- extract_regulated_sets(list_of_degs, names_degs, fc_thresh=fc_thresh)
# attach the results to the DEG data frame for each condition, to be used later
DaKD_deg$reg <- regs$daRNAi
DaOE_deg$reg <- regs$da
DaDaOE_deg$reg <- regs$`da:da`
ScOE_deg$reg <- regs$scute

regs[500:504,]
```

Now we need to turn this into logicals:
```{r extract_up|down}
upreg   <- get_deg_logical(regs,'up')
downreg <- get_deg_logical(regs,'down')
```


# 3.2 Plot diagrams


Using instructions from the [vignette](https://cran.r-project.org/web/packages/eulerr/vignettes/introduction.html):
```{r euler_plots, results='hide'}
up_colours <- c(
  brewer.pal(9,'BrBG')[3],   # daRNAi   brown
  brewer.pal(9,'YlOrBr')[4], # daOVEX   yellow/orange
  brewer.pal(9,'Reds')[5],   # dadaOVEX red
  brewer.pal(9,'PuRd')[3]    # scOVEX   violet
)
down_colours <- c(
  brewer.pal(11,'PRGn')[8],  # daRNAi   green
  brewer.pal(11,'RdYlBu')[8], # daOVEX   pale blue
  brewer.pal(11,'RdBu')[9],  # dadaOVEX blue
  brewer.pal(9,'Purples')[5]     # scOVEX   purple
)
upfit <- euler(upreg, loss = 'region', loss_aggregator = 'max')
downfit <- euler(downreg, loss = 'region', loss_aggregator = 'max')
upplot <- plot(upfit,
               fill = up_colours,
               quantities = TRUE)
downplot <- plot(downfit,
                 fill = down_colours,
                 quantities = TRUE)
# save it
pdf(file=paste0(figdir,'/euler_up.pdf'),
     width=10, height=12)
upplot
dev.off()

pdf(file=paste0(figdir,'/euler_down.pdf'),
     width=10, height=12)
downplot
dev.off()

print(upplot)
print(downplot)
```


# 4 RNAseq results visualisation: MA plots


## 4.1 List of marker genes

The aim is to have a list of genes that could serve as cell type _exclusive_ markers. We will only consider a simplified cell type classification: ISC, EB, EE, EC of the posterior midgut. They must be either expressed exclusively in a cell type (e.g. _Delta_), or functionally promote one cell type only (e.g. _phyllopod_). We start with a curated list  and the markers defined by the Fly Cell Atlas consortium. Then we will remove redundancy.

- First make our list.

```{r load_GOI, message=FALSE}
genes_of_interest <- read_excel('resources/genes_of_interest.xlsx', sheet='core') %>%
  dplyr::select(gene_symbol, bonafide_celltype) %>%
  filter(bonafide_celltype != 'NA') %>%
  rename(celltype = bonafide_celltype) %>%
  group_by(celltype) %>%
  summarise(gene_symbol=paste(gene_symbol, collapse=', '))
```






















- Then the FCA markers.

```{r loadFCA, message=FALSE}
markers <- read_xlsx(file.path(getwd(), 'resources', 'science.abk2432_table s2.xlsx')) %>%
  # get only the gut epithelium/muscle cell types
  filter(Tissue=='gut' | `Annotation...1`=='enteroendocrine cell') %>%
  # remove a hybrid/transitional cell type
  filter(!str_detect(`Annotation...1`, 'differentiating')) %>%
  # remove muscle/artefact cells
  filter(!str_detect(`Broad annotation`, 'muscle|artefact')) %>%
  # get just the columns with marker names
  dplyr::select(contains(c('Annotation...1', 'Marker'))) %>%
  # join all markers in a new column
  unite(gene_symbol, contains('Markers'), sep = ', ')
names(markers)[[1]] <- 'celltype'
```

- An merge the two:

```{r merge_markers, message=FALSE}
missd.spelld <- c(`wntD, `='', `LManIV, `='', `CG31269, `='', `CG43208, `='', `orcokinin B, `='',
                   AstB='Mip', poxn='Poxn')
markers <- rbind(genes_of_interest, markers) %>%
  # merge same cell type markers
  group_by(celltype) %>%
  summarise(gene_symbol=paste(gene_symbol, collapse=', ')) %>%
  # correct difference in synonym/capitalisation with DEG data...
  # ... while all gene symbols are in one string
  mutate(gene_symbol = str_replace_all(gene_symbol, missd.spelld)) %>%
  # make each gene name an independent string
  rowwise() %>% mutate(markers = list(unique(str_split(gene_symbol, ', ')[[1]])))
head(markers)
```

- Filter out the genes that do not have unique mapping to a cell type.

```{r filter_markers, message=FALSE}
# filter by exclusivity, but keeping what is common to all enterocytes
## first identify all EC markers
ECmarkers <- markers %>%
  filter(str_detect(celltype, 'enterocyte')) %>%
  dplyr::select(markers) %>%
  unlist(use.names = FALSE) %>%
  unique() %>% as.list()
## then non-EC cell markers, with all their repetitions!!
nonECmarkers <- markers %>%
  filter(!str_detect(celltype, 'enterocyte')) %>%
  dplyr::select(markers) %>%
  unlist(use.names = FALSE) %>% as.list()
## get all markers with one occurrence (≠unique!)
repmarkers <- unlist(c(ECmarkers, nonECmarkers))
reps <- rle(sort(repmarkers))
xmarkers <- reps$values[reps$lengths==1]
## apply criterion of exclusivity
exclude <- function(x) x[x %in% xmarkers]
markers <- markers %>%
    mutate(exclusive = list(exclude(markers)))
pmgcelltypes <- c('intestinal stem cell', 'enteroblast', 'enteroendocrine cell', 'enterocyte of posterior adult midgut epithelium')
pmg.markers <- markers %>%
  filter(celltype %in% pmgcelltypes) %>%
  dplyr::select(c(1,4)) %>%
  rename(xmarkers = exclusive)
```

- Add cell type colours HEX codes

```{r colour_markers}
write_xlsx(unnest_longer(pmg.markers, xmarkers) %>%
    rename(gene_symbol = xmarkers),
    path = 'output/preliminary_Table S4.xlsx')
cellcolours <- c('EB'='#56B2E9', 'EC'='#009E73', 'EE'='#CC79A7', 'ISC'='#D55E00')
pmg.markers$cellcolour <- cellcolours
```


## 2 MA plots using `ggpubr::ggmaplot` wrappers


MA plot for *daughterless* knockdown:
```{r ggmaplot2_da_knockdown}
repulsion <- list(box.padding = 0.1,
                  point.padding = 0.8,
                  nudge_x = 3,
                  nudge_y = 2,
                  force = 1,
                  force_pull = 0,
                  seed.up = 42,
                  seed.dn = 50,
                  xlims.up = c(14, NA),
                  ylims.up = c(5, NA),
                  xlims.dn = c(18, NA),
                  ylims.dn = c(NA, -1))

ggmaplot2(deg = DaKD_deg,
          fc_thresh = fc_thresh,
          markers = pmg.markers,
          md_label = '*esg > da^RNAi^*',
          repulsion = repulsion)
```

Now with cell type-specific colours:
```{r ggmaplot3_da_kd}
repulsion <- list(box.padding = 0.08,
                  point.padding = 0.1,
                  nudge_x = 0,
                  nudge_y = 0,
                  force = 1,
                  force_pull = -0.004,
                  seed.up = 42,
                  seed.dn = 5,
                  xlims.up = c(12, NA),
                  ylims.up = c(5.5, NA),
                  xlims.dn = c(16, NA),
                  ylims.dn = c(NA, -1))

ggmaplot3(deg = DaKD_deg,
          fc_thresh = fc_thresh,
          markers = pmg.markers, 
          md_label = '*esg > da^RNAi^*',
          repulsion = repulsion)


suppressMessages(ggsave('MA_daKD.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```
  
```{r ggmaplot3_da_overexpression}
repulsion <- list(box.padding = 0.1,
                  point.padding = 0.1,
                  force = 1.5,
                  force_pull = -0.003,
                  nudge_x.up = 0,
                  nudge_y.up = 0,
                  nudge_x.dn = 0,
                  nudge_y.dn = 0,
                  seed.up = 42,
                  seed.dn = 5,
                  xlims.up = c(12, NA),
                  ylims.up = c(5, NA),
                  xlims.dn = c(16, NA),
                  ylims.dn = c(NA, -1))

ggmaplot3(deg = DaOE_deg,
          fc_thresh = fc_thresh,
          markers = pmg.markers, 
          md_label = '*esg > da*',
          repulsion = repulsion)

suppressMessages(ggsave('MA_daOE.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```

To the Daughterless homodimer overexpression:
```{r ggmaplot2_da:da_overexpression}
repulsion <- list(box.padding = 0.01,
                  point.padding = 0.1,
                  nudge_x = 0,
                  nudge_y = 0,
                  force = 1,
                  force_pull = -0.004,
                  seed.up = 42,
                  seed.dn = 5,
                  xlims.up = c(16, NA),
                  ylims.up = c(3, NA),
                  xlims.dn = c(16, NA),
                  ylims.dn = c(NA, -2))

ggmaplot3(deg = DaDaOE_deg,
          fc_thresh = fc_thresh,
          markers = pmg.markers, 
          md_label = '*esg > da:da*',
          repulsion = repulsion)

suppressMessages(ggsave('MA_dadaOE.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```

And, finally, to the *scute* overexpression:
```{r ggmaplot2_sc_overexpression}
repulsion <- list(box.padding = 0.01,
                  point.padding = 0.1,
                  nudge_x = 0,
                  nudge_y = 0,
                  force = 0.7,
                  force_pull = -0.003,
                  seed.up = 42,
                  seed.dn = 5,
                  xlims.up = c(8, NA),
                  ylims.up = c(7, NA),
                  xlims.dn = c(8, NA),
                  ylims.dn = c(NA, -6))

ggmaplot3(deg = ScOE_deg,
          fc_thresh = 2,
          markers = pmg.markers,
          md_label = '*esg > scute*',
          repulsion = repulsion)

suppressMessages(ggsave('MA_ScOE.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```


## 3 Heatmap of individual genes


```{r}
tidy.markers <- pmg.markers %>% unnest_longer(xmarkers) %>%
  rename(gene_symbol = xmarkers)

DaKD_deg$condition <- '*da^RNAi^*'
DaOE_deg$condition <- '*da*'
DaDaOE_deg$condition <- '*da:da*'
ScOE_deg$condition <- '*scute*'
genehm.df <- bind_rows(ScOE_deg, DaDaOE_deg, DaOE_deg, DaKD_deg) %>%
  dplyr::select(ensemblGeneID, gene_symbol, log2FoldChange, padj, condition) %>%
  rename(p.adjust = padj) %>%
  filter(gene_symbol %in% tidy.markers$gene_symbol) %>%
  merge(., tidy.markers, by = 'gene_symbol') %>%
  mutate(condition = factor(condition, levels=c("*scute*", "*da:da*", "*da*", "*da^RNAi^*")))
```


```{r, fig.width=20}
p <- layer.heatmaph(genehm.df, arr='celltype')
p + geom_hline(aes(yintercept=3.5), linewidth = 0.5) +
  theme(plot.margin = margin(r = 25))
```

```{r, fig.width=20}
p <- layer.heatmaph(genehm.df, cluster=TRUE)
p + geom_hline(aes(yintercept=3.5), linewidth = 0.5) +
  theme(plot.margin = margin(r = 25))
```


```{r, fig.height = 20}
p <- layer.heatmapv(genehm.df, cluster=TRUE)
p + geom_vline(aes(xintercept=3.5), linewidth = 0.5)
suppressMessages(ggsave('marker_heatmap.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```

```{r}
Espl.df <- bind_rows(ScOE_deg, DaDaOE_deg, DaOE_deg, DaKD_deg) %>%
  dplyr::select(ensemblGeneID, gene_symbol, log2FoldChange, padj, condition) %>%
  rename(p.adjust = padj) %>%
  filter(grepl("^E\\(spl\\)", gene_symbol)) %>%
  mutate(condition = factor(condition, levels=c("*scute*", "*da:da*", "*da*", "*da^RNAi^*"))) %>%
  mutate(cellcolour = "#000000")

p <- layer.heatmapv(Espl.df, cluster=TRUE)
p + geom_vline(aes(xintercept=3.5), linewidth = 0.5)
suppressMessages(ggsave('Esplit_heatmap.pdf', plot = last_plot(), device = 'pdf',
                        path = figdir, dpi = 300))
```

```{r}
# License: https://creativecommons.org/publicdomain/zero/1.0/
# xkcd colour name list
cloudy blue	#acc2d9	
dark pastel green	#56ae57	
dust	#b2996e	
electric lime	#a8ff04	
fresh green	#69d84f	
light eggplant	#894585	
nasty green	#70b23f	
really light blue	#d4ffff	
tea	#65ab7c	
warm purple	#952e8f	
yellowish tan	#fcfc81	
cement	#a5a391	
dark grass green	#388004	
dusty teal	#4c9085	
grey teal	#5e9b8a	
macaroni and cheese	#efb435	
pinkish tan	#d99b82	
spruce	#0a5f38	
strong blue	#0c06f7	
toxic green	#61de2a	
windows blue	#3778bf	
blue blue	#2242c7	
blue with a hint of purple	#533cc6	
booger	#9bb53c	
bright sea green	#05ffa6	
dark green blue	#1f6357	
deep turquoise	#017374	
green teal	#0cb577	
strong pink	#ff0789	
bland	#afa88b	
deep aqua	#08787f	
lavender pink	#dd85d7	
light moss green	#a6c875	
light seafoam green	#a7ffb5	
olive yellow	#c2b709	
pig pink	#e78ea5	
deep lilac	#966ebd	
desert	#ccad60	
dusty lavender	#ac86a8	
purpley grey	#947e94	
purply	#983fb2	
candy pink	#ff63e9	
light pastel green	#b2fba5	
boring green	#63b365	
kiwi green	#8ee53f	
light grey green	#b7e1a1	
orange pink	#ff6f52	
tea green	#bdf8a3	
very light brown	#d3b683	
egg shell	#fffcc4	
eggplant purple	#430541	
powder pink	#ffb2d0	
reddish grey	#997570	
baby shit brown	#ad900d	
liliac	#c48efd	
stormy blue	#507b9c	
ugly brown	#7d7103	
custard	#fffd78	
darkish pink	#da467d	
deep brown	#410200	
greenish beige	#c9d179	
manilla	#fffa86	
off blue	#5684ae	
battleship grey	#6b7c85	
browny green	#6f6c0a	
bruise	#7e4071	
kelley green	#009337	
sickly yellow	#d0e429	
sunny yellow	#fff917	
azul	#1d5dec	
darkgreen	#054907	
green/yellow	#b5ce08	
lichen	#8fb67b	
light light green	#c8ffb0	
pale gold	#fdde6c	
sun yellow	#ffdf22	
tan green	#a9be70	
burple	#6832e3	
butterscotch	#fdb147	
toupe	#c7ac7d	
dark cream	#fff39a	
indian red	#850e04	
light lavendar	#efc0fe	
poison green	#40fd14	
baby puke green	#b6c406	
bright yellow green	#9dff00	
charcoal grey	#3c4142	
squash	#f2ab15	
cinnamon	#ac4f06	
light pea green	#c4fe82	
radioactive green	#2cfa1f	
raw sienna	#9a6200	
baby purple	#ca9bf7	
cocoa	#875f42	
light royal blue	#3a2efe	
orangeish	#fd8d49	
rust brown	#8b3103	
sand brown	#cba560	
swamp	#698339	
tealish green	#0cdc73	
burnt siena	#b75203	
camo	#7f8f4e	
dusk blue	#26538d	
fern	#63a950	
old rose	#c87f89	
pale light green	#b1fc99	
peachy pink	#ff9a8a	
rosy pink	#f6688e	
light bluish green	#76fda8	
light bright green	#53fe5c	
light neon green	#4efd54	
light seafoam	#a0febf	
tiffany blue	#7bf2da	
washed out green	#bcf5a6	
browny orange	#ca6b02	
nice blue	#107ab0	
sapphire	#2138ab	
greyish teal	#719f91	
orangey yellow	#fdb915	
parchment	#fefcaf	
straw	#fcf679	
very dark brown	#1d0200	
terracota	#cb6843	
ugly blue	#31668a	
clear blue	#247afd	
creme	#ffffb6	
foam green	#90fda9	
grey/green	#86a17d	
light gold	#fddc5c	
seafoam blue	#78d1b6	
topaz	#13bbaf	
violet pink	#fb5ffc	
wintergreen	#20f986	
yellow tan	#ffe36e	
dark fuchsia	#9d0759	
indigo blue	#3a18b1	
light yellowish green	#c2ff89	
pale magenta	#d767ad	
rich purple	#720058	
sunflower yellow	#ffda03	
green/blue	#01c08d	
leather	#ac7434	
racing green	#014600	
vivid purple	#9900fa	
dark royal blue	#02066f	
hazel	#8e7618	
muted pink	#d1768f	
booger green	#96b403	
canary	#fdff63	
cool grey	#95a3a6	
dark taupe	#7f684e	
darkish purple	#751973	
true green	#089404	
coral pink	#ff6163	
dark sage	#598556	
dark slate blue	#214761	
flat blue	#3c73a8	
mushroom	#ba9e88	
rich blue	#021bf9	
dirty purple	#734a65	
greenblue	#23c48b	
icky green	#8fae22	
light khaki	#e6f2a2	
warm blue	#4b57db	
dark hot pink	#d90166	
deep sea blue	#015482	
carmine	#9d0216	
dark yellow green	#728f02	
pale peach	#ffe5ad	
plum purple	#4e0550	
golden rod	#f9bc08	
neon red	#ff073a	
old pink	#c77986	
very pale blue	#d6fffe	
blood orange	#fe4b03	
grapefruit	#fd5956	
sand yellow	#fce166	
clay brown	#b2713d	
dark blue grey	#1f3b4d	
flat green	#699d4c	
light green blue	#56fca2	
warm pink	#fb5581	
dodger blue	#3e82fc	
gross green	#a0bf16	
ice	#d6fffa	
metallic blue	#4f738e	
pale salmon	#ffb19a	
sap green	#5c8b15	
algae	#54ac68	
bluey grey	#89a0b0	
greeny grey	#7ea07a	
highlighter green	#1bfc06	
light light blue	#cafffb	
light mint	#b6ffbb	
raw umber	#a75e09	
vivid blue	#152eff	
deep lavender	#8d5eb7	
dull teal	#5f9e8f	
light greenish blue	#63f7b4	
mud green	#606602	
pinky	#fc86aa	
red wine	#8c0034	
shit green	#758000	
tan brown	#ab7e4c	
darkblue	#030764	
rosa	#fe86a4	
lipstick	#d5174e	
pale mauve	#fed0fc	
claret	#680018	
dandelion	#fedf08	
orangered	#fe420f	
poop green	#6f7c00	
ruby	#ca0147	
dark	#1b2431	
greenish turquoise	#00fbb0	
pastel red	#db5856	
piss yellow	#ddd618	
bright cyan	#41fdfe	
dark coral	#cf524e	
algae green	#21c36f	
darkish red	#a90308	
reddy brown	#6e1005	
blush pink	#fe828c	
camouflage green	#4b6113	
lawn green	#4da409	
putty	#beae8a	
vibrant blue	#0339f8	
dark sand	#a88f59	
purple/blue	#5d21d0	
saffron	#feb209	
twilight	#4e518b	
warm brown	#964e02	
bluegrey	#85a3b2	
bubble gum pink	#ff69af	
duck egg blue	#c3fbf4	
greenish cyan	#2afeb7	
petrol	#005f6a	
royal	#0c1793	
butter	#ffff81	
dusty orange	#f0833a	
off yellow	#f1f33f	
pale olive green	#b1d27b	
orangish	#fc824a	
leaf	#71aa34	
light blue grey	#b7c9e2	
dried blood	#4b0101	
lightish purple	#a552e6	
rusty red	#af2f0d	
lavender blue	#8b88f8	
light grass green	#9af764	
light mint green	#a6fbb2	
sunflower	#ffc512	
velvet	#750851	
brick orange	#c14a09	
lightish red	#fe2f4a	
pure blue	#0203e2	
twilight blue	#0a437a	
violet red	#a50055	
yellowy brown	#ae8b0c	
carnation	#fd798f	
muddy yellow	#bfac05	
dark seafoam green	#3eaf76	
deep rose	#c74767	
dusty red	#b9484e	
grey/blue	#647d8e	
lemon lime	#bffe28	
purple/pink	#d725de	
brown yellow	#b29705	
purple brown	#673a3f	
wisteria	#a87dc2	
banana yellow	#fafe4b	
lipstick red	#c0022f	
water blue	#0e87cc	
brown grey	#8d8468	
vibrant purple	#ad03de	
baby green	#8cff9e	
barf green	#94ac02	
eggshell blue	#c4fff7	
sandy yellow	#fdee73	
cool green	#33b864	
pale	#fff9d0	
blue/grey	#758da3	
hot magenta	#f504c9	
greyblue	#77a1b5	
purpley	#8756e4	
baby shit green	#889717	
brownish pink	#c27e79	
dark aquamarine	#017371	
diarrhea	#9f8303	
light mustard	#f7d560	
pale sky blue	#bdf6fe	
turtle green	#75b84f	
bright olive	#9cbb04	
dark grey blue	#29465b	
greeny brown	#696006	
lemon green	#adf802	
light periwinkle	#c1c6fc	
seaweed green	#35ad6b	
sunshine yellow	#fffd37	
ugly purple	#a442a0	
medium pink	#f36196	
puke brown	#947706	
very light pink	#fff4f2	
viridian	#1e9167	
bile	#b5c306	
faded yellow	#feff7f	
very pale green	#cffdbc	
vibrant green	#0add08	
bright lime	#87fd05	
spearmint	#1ef876	
light aquamarine	#7bfdc7	
light sage	#bcecac	
yellowgreen	#bbf90f	
baby poo	#ab9004	
dark seafoam	#1fb57a	
deep teal	#00555a	
heather	#a484ac	
rust orange	#c45508	
dirty blue	#3f829d	
fern green	#548d44	
bright lilac	#c95efb	
weird green	#3ae57f	
peacock blue	#016795	
avocado green	#87a922	
faded orange	#f0944d	
grape purple	#5d1451	
hot green	#25ff29	
lime yellow	#d0fe1d	
mango	#ffa62b	
shamrock	#01b44c	
bubblegum	#ff6cb5	
purplish brown	#6b4247	
vomit yellow	#c7c10c	
pale cyan	#b7fffa	
key lime	#aeff6e	
tomato red	#ec2d01	
lightgreen	#76ff7b	
merlot	#730039	
night blue	#040348	
purpleish pink	#df4ec8	
apple	#6ecb3c	
baby poop green	#8f9805	
green apple	#5edc1f	
heliotrope	#d94ff5	
yellow/green	#c8fd3d	
almost black	#070d0d	
cool blue	#4984b8	
leafy green	#51b73b	
mustard brown	#ac7e04	
dusk	#4e5481	
dull brown	#876e4b	
frog green	#58bc08	
vivid green	#2fef10	
bright light green	#2dfe54	
fluro green	#0aff02	
kiwi	#9cef43	
seaweed	#18d17b	
navy green	#35530a	
ultramarine blue	#1805db	
iris	#6258c4	
pastel orange	#ff964f	
yellowish orange	#ffab0f	
perrywinkle	#8f8ce7	
tealish	#24bca8	
dark plum	#3f012c	
pear	#cbf85f	
pinkish orange	#ff724c	
midnight purple	#280137	
light urple	#b36ff6	
dark mint	#48c072	
greenish tan	#bccb7a	
light burgundy	#a8415b	
turquoise blue	#06b1c4	
ugly pink	#cd7584	
sandy	#f1da7a	
electric pink	#ff0490	
muted purple	#805b87	
mid green	#50a747	
greyish	#a8a495	
neon yellow	#cfff04	
banana	#ffff7e	
carnation pink	#ff7fa7	
tomato	#ef4026	
sea	#3c9992	
muddy brown	#886806	
turquoise green	#04f489	
buff	#fef69e	
fawn	#cfaf7b	
muted blue	#3b719f	
pale rose	#fdc1c5	
dark mint green	#20c073	
amethyst	#9b5fc0	
blue/green	#0f9b8e	
chestnut	#742802	
sick green	#9db92c	
pea	#a4bf20	
rusty orange	#cd5909	
stone	#ada587	
rose red	#be013c	
pale aqua	#b8ffeb	
deep orange	#dc4d01	
earth	#a2653e	
mossy green	#638b27	
grassy green	#419c03	
pale lime green	#b1ff65	
light grey blue	#9dbcd4	
pale grey	#fdfdfe	
asparagus	#77ab56	
blueberry	#464196	
purple red	#990147	
pale lime	#befd73	
greenish teal	#32bf84	
caramel	#af6f09	
deep magenta	#a0025c	
light peach	#ffd8b1	
milk chocolate	#7f4e1e	
ocher	#bf9b0c	
off green	#6ba353	
purply pink	#f075e6	
lightblue	#7bc8f6	
dusky blue	#475f94	
golden	#f5bf03	
light beige	#fffeb6	
butter yellow	#fffd74	
dusky purple	#895b7b	
french blue	#436bad	
ugly yellow	#d0c101	
greeny yellow	#c6f808	
orangish red	#f43605	
shamrock green	#02c14d	
orangish brown	#b25f03	
tree green	#2a7e19	
deep violet	#490648	
gunmetal	#536267	
blue/purple	#5a06ef	
cherry	#cf0234	
sandy brown	#c4a661	
warm grey	#978a84	
dark indigo	#1f0954	
midnight	#03012d	
bluey green	#2bb179	
grey pink	#c3909b	
soft purple	#a66fb5	
blood	#770001	
brown red	#922b05	
medium grey	#7d7f7c	
berry	#990f4b	
poo	#8f7303	
purpley pink	#c83cb9	
light salmon	#fea993	
snot	#acbb0d	
easter purple	#c071fe	
light yellow green	#ccfd7f	
dark navy blue	#00022e	
drab	#828344	
light rose	#ffc5cb	
rouge	#ab1239	
purplish red	#b0054b	
slime green	#99cc04	
baby poop	#937c00	
irish green	#019529	
pink/purple	#ef1de7	
dark navy	#000435	
greeny blue	#42b395	
light plum	#9d5783	
pinkish grey	#c8aca9	
dirty orange	#c87606	
rust red	#aa2704	
pale lilac	#e4cbff	
orangey red	#fa4224	
primary blue	#0804f9	
kermit green	#5cb200	
brownish purple	#76424e	
murky green	#6c7a0e	
wheat	#fbdd7e	
very dark purple	#2a0134	
bottle green	#044a05	
watermelon	#fd4659	
deep sky blue	#0d75f8	
fire engine red	#fe0002	
yellow ochre	#cb9d06	
pumpkin orange	#fb7d07	
pale olive	#b9cc81	
light lilac	#edc8ff	
lightish green	#61e160	
carolina blue	#8ab8fe	
mulberry	#920a4e	
shocking pink	#fe02a2	
auburn	#9a3001	
bright lime green	#65fe08	
celadon	#befdb7	
pinkish brown	#b17261	
poo brown	#885f01	
bright sky blue	#02ccfe	
celery	#c1fd95	
dirt brown	#836539	
strawberry	#fb2943	
dark lime	#84b701	
copper	#b66325	
medium brown	#7f5112	
muted green	#5fa052	
robin's egg	#6dedfd	
bright aqua	#0bf9ea	
bright lavender	#c760ff	
ivory	#ffffcb	
very light purple	#f6cefc	
light navy	#155084	
pink red	#f5054f	
olive brown	#645403	
poop brown	#7a5901	
mustard green	#a8b504	
ocean green	#3d9973	
very dark blue	#000133	
dusty green	#76a973	
light navy blue	#2e5a88	
minty green	#0bf77d	
adobe	#bd6c48	
barney	#ac1db8	
jade green	#2baf6a	
bright light blue	#26f7fd	
light lime	#aefd6c	
dark khaki	#9b8f55	
orange yellow	#ffad01	
ocre	#c69c04	
maize	#f4d054	
faded pink	#de9dac	
british racing green	#05480d	
sandstone	#c9ae74	
mud brown	#60460f	
light sea green	#98f6b0	
robin egg blue	#8af1fe	
aqua marine	#2ee8bb	
dark sea green	#11875d	
soft pink	#fdb0c0	
orangey brown	#b16002	
cherry red	#f7022a	
burnt yellow	#d5ab09	
brownish grey	#86775f	
camel	#c69f59	
purplish grey	#7a687f	
marine	#042e60	
greyish pink	#c88d94	
pale turquoise	#a5fbd5	
pastel yellow	#fffe71	
bluey purple	#6241c7	
canary yellow	#fffe40	
faded red	#d3494e	
sepia	#985e2b	
coffee	#a6814c	
bright magenta	#ff08e8	
mocha	#9d7651	
ecru	#feffca	
purpleish	#98568d	
cranberry	#9e003a	
darkish green	#287c37	
brown orange	#b96902	
dusky rose	#ba6873	
melon	#ff7855	
sickly green	#94b21c	
silver	#c5c9c7	
purply blue	#661aee	
purpleish blue	#6140ef	
hospital green	#9be5aa	
shit brown	#7b5804	
mid blue	#276ab3	
amber	#feb308	
easter green	#8cfd7e	
soft blue	#6488ea	
cerulean blue	#056eee	
golden brown	#b27a01	
bright turquoise	#0ffef9	
red pink	#fa2a55	
red purple	#820747	
greyish brown	#7a6a4f	
vermillion	#f4320c	
russet	#a13905	
steel grey	#6f828a	
lighter purple	#a55af4	
bright violet	#ad0afd	
prussian blue	#004577	
slate green	#658d6d	
dirty pink	#ca7b80	
dark blue green	#005249	
pine	#2b5d34	
yellowy green	#bff128	
dark gold	#b59410	
bluish	#2976bb	
darkish blue	#014182	
dull red	#bb3f3f	
pinky red	#fc2647	
bronze	#a87900	
pale teal	#82cbb2	
military green	#667c3e	
barbie pink	#fe46a5	
bubblegum pink	#fe83cc	
pea soup green	#94a617	
dark mustard	#a88905	
shit	#7f5f00	
medium purple	#9e43a2	
very dark green	#062e03	
dirt	#8a6e45	
dusky pink	#cc7a8b	
red violet	#9e0168	
lemon yellow	#fdff38	
pistachio	#c0fa8b	
dull yellow	#eedc5b	
dark lime green	#7ebd01	
denim blue	#3b5b92	
teal blue	#01889f	
lightish blue	#3d7afd	
purpley blue	#5f34e7	
light indigo	#6d5acf	
swamp green	#748500	
brown green	#706c11	
dark maroon	#3c0008	
hot purple	#cb00f5	
dark forest green	#002d04	
faded blue	#658cbb	
drab green	#749551	
light lime green	#b9ff66	
snot green	#9dc100	
yellowish	#faee66	
light blue green	#7efbb3	
bordeaux	#7b002c	
light mauve	#c292a1	
ocean	#017b92	
marigold	#fcc006	
muddy green	#657432	
dull orange	#d8863b	
steel	#738595	
electric purple	#aa23ff	
fluorescent green	#08ff08	
yellowish brown	#9b7a01	
blush	#f29e8e	
soft green	#6fc276	
bright orange	#ff5b00	
lemon	#fdff52	
purple grey	#866f85	
acid green	#8ffe09	
pale lavender	#eecffe	
violet blue	#510ac9	
light forest green	#4f9153	
burnt red	#9f2305	
khaki green	#728639	
cerise	#de0c62	
faded purple	#916e99	
apricot	#ffb16d	
dark olive green	#3c4d03	
grey brown	#7f7053	
green grey	#77926f	
true blue	#010fcc	
pale violet	#ceaefa	
periwinkle blue	#8f99fb	
light sky blue	#c6fcff	
blurple	#5539cc	
green brown	#544e03	
bluegreen	#017a79	
bright teal	#01f9c6	
brownish yellow	#c9b003	
pea soup	#929901	
forest	#0b5509	
barney purple	#a00498	
ultramarine	#2000b1	
purplish	#94568c	
puke yellow	#c2be0e	
bluish grey	#748b97	
dark periwinkle	#665fd1	
dark lilac	#9c6da5	
reddish	#c44240	
light maroon	#a24857	
dusty purple	#825f87	
terra cotta	#c9643b	
avocado	#90b134	
marine blue	#01386a	
teal green	#25a36f	
slate grey	#59656d	
lighter green	#75fd63	
electric green	#21fc0d	
dusty blue	#5a86ad	
golden yellow	#fec615	
bright yellow	#fffd01	
light lavender	#dfc5fe	
umber	#b26400	
poop	#7f5e00	
dark peach	#de7e5d	
jungle green	#048243	
eggshell	#ffffd4	
denim	#3b638c	
yellow brown	#b79400	
dull purple	#84597e	
chocolate brown	#411900	
wine red	#7b0323	
neon blue	#04d9ff	
dirty green	#667e2c	
light tan	#fbeeac	
ice blue	#d7fffe	
cadet blue	#4e7496	
dark mauve	#874c62	
very light blue	#d5ffff	
grey purple	#826d8c	
pastel pink	#ffbacd	
very light green	#d1ffbd	
dark sky blue	#448ee4	
evergreen	#05472a	
dull pink	#d5869d	
aubergine	#3d0734	
mahogany	#4a0100	
reddish orange	#f8481c	
deep green	#02590f	
vomit green	#89a203	
purple pink	#e03fd8	
dusty pink	#d58a94	
faded green	#7bb274	
camo green	#526525	
pinky purple	#c94cbe	
pink purple	#db4bda	
brownish red	#9e3623	
dark rose	#b5485d	
mud	#735c12	
brownish	#9c6d57	
emerald green	#028f1e	
pale brown	#b1916e	
dull blue	#49759c	
burnt umber	#a0450e	
medium green	#39ad48	
clay	#b66a50	
light aqua	#8cffdb	
light olive green	#a4be5c	
brownish orange	#cb7723	
dark aqua	#05696b	
purplish pink	#ce5dae	
dark salmon	#c85a53	
greenish grey	#96ae8d	
jade	#1fa774	
ugly green	#7a9703	
dark beige	#ac9362	
emerald	#01a049	
pale red	#d9544d	
light magenta	#fa5ff7	
sky	#82cafc	
light cyan	#acfffc	
yellow orange	#fcb001	
reddish purple	#910951	
reddish pink	#fe2c54	
orchid	#c875c4	
dirty yellow	#cdc50a	
orange red	#fd411e	
deep red	#9a0200	
orange brown	#be6400	
cobalt blue	#030aa7	
neon pink	#fe019a	
rose pink	#f7879a	
greyish purple	#887191	
raspberry	#b00149	
aqua green	#12e193	
salmon pink	#fe7b7c	
tangerine	#ff9408	
brownish green	#6a6e09	
red brown	#8b2e16	
greenish brown	#696112	
pumpkin	#e17701	
pine green	#0a481e	
charcoal	#343837	
baby pink	#ffb7ce	
cornflower	#6a79f7	
blue violet	#5d06e9	
chocolate	#3d1c02	
greyish green	#82a67d	
scarlet	#be0119	
green yellow	#c9ff27	
dark olive	#373e02	
sienna	#a9561e	
pastel purple	#caa0ff	
terracotta	#ca6641	
aqua blue	#02d8e9	
sage green	#88b378	
blood red	#980002	
deep pink	#cb0162	
grass	#5cac2d	
moss	#769958	
pastel blue	#a2bffe	
bluish green	#10a674	
green blue	#06b48b	
dark tan	#af884a	
greenish blue	#0b8b87	
pale orange	#ffa756	
vomit	#a2a415	
forrest green	#154406	
dark lavender	#856798	
dark violet	#34013f	
purple blue	#632de9	
dark cyan	#0a888a	
olive drab	#6f7632	
pinkish	#d46a7e	
cobalt	#1e488f	
neon purple	#bc13fe	
light turquoise	#7ef4cc	
apple green	#76cd26	
dull green	#74a662	
wine	#80013f	
powder blue	#b1d1fc	
off white	#ffffe4	
electric blue	#0652ff	
dark turquoise	#045c5a	
blue purple	#5729ce	
azure	#069af3	
bright red	#ff000d
pinkish red	#f10c45	
cornflower blue	#5170d7	
light olive	#acbf69	
grape	#6c3461	
greyish blue	#5e819d	
purplish blue	#601ef9	
yellowish green	#b0dd16	
greenish yellow	#cdfd02	
medium blue	#2c6fbb	
dusty rose	#c0737a	
light violet	#d6b4fc	
midnight blue	#020035	
bluish purple	#703be7	
red orange	#fd3c06	
dark magenta	#960056	
greenish	#40a368	
ocean blue	#03719c	
coral	#fc5a50	
cream	#ffffc2	
reddish brown	#7f2b0a	
burnt sienna	#b04e0f	
brick	#a03623	
sage	#87ae73	
grey green	#789b73	
white	#ffffff	
robin's egg blue	#98eff9	
moss green	#658b38	
steel blue	#5a7d9a	
eggplant	#380835	
light yellow	#fffe7a	
leaf green	#5ca904	
light grey	#d8dcd6	
puke	#a5a502	
pinkish purple	#d648d7	
sea blue	#047495	
pale purple	#b790d4	
slate blue	#5b7c99	
blue grey	#607c8e	
hunter green	#0b4008	
fuchsia	#ed0dd9	
crimson	#8c000f	
pale yellow	#ffff84	
ochre	#bf9005	
mustard yellow	#d2bd0a	
light red	#ff474c	
cerulean	#0485d1	
pale pink	#ffcfdc	
deep blue	#040273	
rust	#a83c09	
light teal	#90e4c1	
slate	#516572	
goldenrod	#fac205	
dark yellow	#d5b60a	
dark grey	#363737	
army green	#4b5d16	
grey blue	#6b8ba4	
seafoam	#80f9ad	
puce	#a57e52	
spring green	#a9f971	
dark orange	#c65102	
sand	#e2ca76	
pastel green	#b0ff9d	
mint	#9ffeb0	
light orange	#fdaa48	
bright pink	#fe01b1	
chartreuse	#c1f80a	
deep purple	#36013f	
dark brown	#341c02	
taupe	#b9a281	
pea green	#8eab12	
puke green	#9aae07	
kelly green	#02ab2e	
seafoam green	#7af9ab	
blue green	#137e6d	
khaki	#aaa662	
burgundy	#610023	
dark teal	#014d4e	
brick red	#8f1402	
royal purple	#4b006e	
plum	#580f41	
mint green	#8fff9f	
gold	#dbb40c	
baby blue	#a2cffe	
yellow green	#c0fb2d	
bright purple	#be03fd	
dark red	#840000	
pale blue	#d0fefe	
grass green	#3f9b0b	
navy	#01153e	
aquamarine	#04d8b2	
burnt orange	#c04e01	
neon green	#0cff0c	
bright blue	#0165fc	
rose	#cf6275	
light pink	#ffd1df	
mustard	#ceb301	
indigo	#380282	
lime	#aaff32	
sea green	#53fca1	
periwinkle	#8e82fe	
dark pink	#cb416b	
olive green	#677a04	
peach	#ffb07c	
pale green	#c7fdb5	
light brown	#ad8150	
hot pink	#ff028d	
black	#000000	
lilac	#cea2fd	
navy blue	#001146	
royal blue	#0504aa	
beige	#e6daa6	
salmon	#ff796c	
olive	#6e750e	
maroon	#650021	
bright green	#01ff07	
dark purple	#35063e	
mauve	#ae7181	
forest green	#06470c	
aqua	#13eac9	
cyan	#00ffff	
tan	#d1b26f	
dark blue	#00035b	
lavender	#c79fef	
turquoise	#06c2ac	
dark green	#033500	
violet	#9a0eea	
light purple	#bf77f6	
lime green	#89fe05	
grey	#929591	
sky blue	#75bbfd	
yellow	#ffff14	
magenta	#c20078	
light green	#96f97b	
orange	#f97306	
teal	#029386	
light blue	#95d0fc	
red	#e50000	
brown	#653700	
pink	#ff81c0	
blue	#0343df	
green	#15b01a	
purple	#7e1e9c	

```