---
title: "Visualisation of DESeq2 results with sets"
description: "DEG analysis based on DESeq2 and GSEA"
principal investigator: "Joaquín de Navascués"
researcher: "Aleix Puig, modified by Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: readable
    df_print: paged
---
```{r set-publication-theme, echo=FALSE, cache=FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
```

```{r setup, echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 5, fig.width = 8.5, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path='integration/figures/', warning=FALSE, message=FALSE)
```

**Libraries needed:**
```{r load_libraries, warning=FALSE, echo=FALSE}
if (!require("librarian")) install.packages("librarian")
# data
librarian::shelf(dplyr, purrr, stringr, reshape2, biomaRt, DescTools, santoku)
# graphics
librarian::shelf(UpSetR, ggVennDiagram, ggvenn, eulerr, RColorBrewer,
                 ggthemes, ggtext, cetcolor)
# convenience
librarian::shelf(here)
setwd(here::here()) # to distinguish from dplyr::here()
```

**Path to definitive images (outside repo):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(),'/')[[1]],-1),
                   paste0(tail(str_split(getwd(),'/')[[1]],1), '_figures')),
                 collapse='/')
dir.create(figdir, showWarnings = FALSE)
```

# RNAseq results visualisation with sets

## 1 Getting ready

This gets us the DGE data from `DESeq2`, identified by FlyBase/Ensembl ID and gene symbol:
```{r load_DEG_data}
# experimental design and labels
targets <- readRDS('output/targets.RDS')
# DEG data
DaDaOE_deg <- readRDS('output/Control_vs_DaDaOE.RDS')
DaKD_deg <- readRDS('output/Control_vs_DaKD.RDS')
DaOE_deg <- readRDS('output/Control_vs_DaOE.RDS')
ScOE_deg <- readRDS('output/Control_vs_ScOE.RDS')
# gene IDs and symbols
ensembl = useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL",
                     dataset="dmelanogaster_gene_ensembl",
                     host = "https://oct2022.archive.ensembl.org")
filters = listFilters(ensembl)       # defines filters for specific queries
attributes = listAttributes(ensembl) # defines features tro show
dlist <- getBM(attributes=c('ensembl_gene_id', 'external_gene_name'), mart = ensembl)
rownames(dlist) <- dlist$ensembl_gene_id
```

Now, for each condition, we will need to get the lists of misregulated genes for a given threshold of fold change:
```{r extract_regulated_sets, warning = FALSE, results=hide}
# extract genes with abs(log2FC)
extract_regulated_sets <- function(list_of_degs, names_degs, fc_thresh=2) {
  # fc_thresh must be a positive number
  # list_of_degs is a list of dataframes
  # names_degs is a list of strings
  # they must have the same length
  reg_lvl <- paste0(' reg@log~2~FC≥', fc_thresh)
  breaks <- brk_manual(c(-fc_thresh, fc_thresh), left_vec = c(FALSE, TRUE))
  regulated_sets <- NULL
  for (l in 1:length(list_of_degs)) {
    degs_na <- dplyr::select(list_of_degs[[l]], c(log2FoldChange, padj, ensemblGeneID))
    # better not pass NAs to kiru
    degs <- na.omit(degs_na)
    # filter by log2FC threshold
    degs$reg <- 
      kiru(
        degs$log2FoldChange,
        breaks = breaks,
        extend = TRUE,
        labels=c("down", "non", "up")
      )
    # filter by p-val
    degs$reg <- as.character(degs$reg)
    x <- 1:nrow(degs)
    degs$reg <- ifelse(degs[x,'padj']<0.5, degs[x,'reg'], 'non')
    # recover NAs as non-regulataed
    added_nas <- degs_na[!(rownames(degs_na) %in% rownames(na.omit(degs_na))),]
    added_nas$reg <- 'non'
    degs <- rbind(degs, added_nas)
    # store reg status
    #rownames(degs) <- degs$ensemblGeneID
    degs[ , names_degs[[l]] ] <- degs$reg
    regulated_sets[[l]] <- dplyr::select(degs, c(ensemblGeneID, names_degs[[l]]) )
  }
  regulated_sets <- purrr::reduce(regulated_sets, full_join, by='ensemblGeneID')
  rownames(regulated_sets) <- regulated_sets$ensemblGeneID
  regulated_sets <- dplyr::select(regulated_sets, -ensemblGeneID)
  return(regulated_sets)
}

list_of_degs <- list(DaKD_deg, DaOE_deg, DaDaOE_deg, ScOE_deg)
names_degs <- list(
  as.character(targets[targets$Condition=='DaKD','condition_md'][[1]]),
  as.character(targets[targets$Condition=='DaOE','condition_md'][[1]]),
  as.character(targets[targets$Condition=='DaDaOE','condition_md'][[1]]),
  as.character(targets[targets$Condition=='ScOE','condition_md'][[1]])
  )
regs <- extract_regulated_sets(list_of_degs, names_degs, fc_thresh=2)
head(regs,2)
```

Now we need to turn this into logicals:
```{r extract_up|down}
get_deg_logical <- function(regs, direction){
  # direction must be 'up' or 'down'
  # regs is the output of `extract_regulated_sets`
  deg_logical_set <- (regs == direction)[
    ( rowSums(regs==direction)!=0 ), 
  ]
  return(deg_logical_set)
}
upreg   <- get_deg_logical(regs,'up')
downreg <- get_deg_logical(regs,'down')
```


## 2 Euler and Venn diagrams

### Euler diagrams using `eulerr`

Using instructions from the [vignette](https://cran.r-project.org/web/packages/eulerr/vignettes/introduction.html):
```{r euler_plots, results=hide}
up_colours <- c(
  brewer.pal(9,'BrBG')[3],   # daRNAi   brown
  brewer.pal(9,'YlOrBr')[4], # daOVEX   yellow/orange
  brewer.pal(9,'Reds')[5],   # dadaOVEX red
  brewer.pal(9,'PuRd')[3]    # scOVEX   violet
)
down_colours <- c(
  brewer.pal(11,'PRGn')[8],  # daRNAi   green
  brewer.pal(11,'RdYlBu')[8], # daOVEX   pale blue
  brewer.pal(11,'RdBu')[9],  # dadaOVEX blue
  brewer.pal(9,'Purples')[5]     # scOVEX   purple
)
upfit <- euler(upreg, loss = 'region', loss_aggregator = 'max')
downfit <- euler(downreg, loss = 'region', loss_aggregator = 'max')
upplot <- plot(upfit,
               fill = up_colours,
               quantities = TRUE)
downplot <- plot(downfit,
                 fill = down_colours,
                 quantities = TRUE)
# save it
pdf(file=paste0(figdir,'/euler_up.pdf'),
     width=10, height=12)
upplot
dev.off()

pdf(file=paste0(figdir,'/euler_down.pdf'),
     width=10, height=12)
downplot
dev.off()

svg(file=paste0(figdir,'/euler_up.svg'),
     width=10, height=12)
upplot
dev.off()

svg(file=paste0(figdir,'/euler_down.svg'),
     width=10, height=12)
downplot
dev.off()

print(upplot)
print(downplot)
```

### Venn diagrams using `ggvenn`

We now need the first column to have the names of the elements in the set, so the `rownames` need become the first column:
```{r prepare_data_for_ggvenn, results=hide}
upreg   <- cbind(data.frame(ensemblGeneID=rownames(upreg)), upreg)
downreg <- cbind(data.frame(ensemblGeneID=rownames(downreg)), downreg)
head(upreg,2)
```

Now both `upreg` and `downreg` are data.frames, which is what `ggvenn` wants as input.

Using instructions from the [vignette](https://github.com/yanlinlin82/ggvenn):
```{r plot_ggvenn_up, fig.width=18, warning=FALSE}
upvenn <- ggplot(upreg,
                 aes(A = `*da^RNAi^*`,
                     B = `*da^OVEX^*`,
                     C = `*da:da^OVEX^*`,
                     D = `*scute^OVEX^*`)) +
  geom_venn(show_percentage = FALSE) +
  theme_void() +
  coord_fixed()
upvenn
```
```{r prepare_data_for_ggvenn, results=hide}
#names(upreg) <- c('ensemblGeneID', 'daKD', 'daOE', 'dadaOE', 'scOE')
names(downreg) <- c('ensemblGeneID', 'daKD', 'daOE', 'dadaOE', 'scOE')
head(downreg,2)
```

```{r plot_ggvenn_down, fig.width=18, warning=FALSE}
downvenn <- ggplot(downreg,
                   aes(A = `daKD`,
                       B = `daOE`,
                       C = `dadaOE`,
                       D = `scOE`)) +
  geom_venn(show_percentage = FALSE) +
  theme_void() +
  coord_fixed()
downvenn
```

This is, however, not very informative, visually. Maybe with `ggVennDiagram`?

### Venn diagrams using `ggVennDiagram`

From the CRAN [readme](https://cran.r-project.org/web/packages/ggVennDiagram/readme/README.html):

```{r}
up <- list(`*da^RNAi^*`    = upreg[upreg$`*da^RNAi^*`,]$ensemblGeneID,
           `*da^OVEX^*`    = upreg[upreg$`*da^OVEX^*`,]$ensemblGeneID,
           `*da:da^OVEX^*` = upreg[upreg$`*da:da^OVEX^*`,]$ensemblGeneID,
           `*scute^OVEX^*` = upreg[upreg$`*scute^OVEX^*`,]$ensemblGeneID)
upvd <- ggVennDiagram(up) +
  scale_fill_gradient(low='white',
                      high = cet_pal(n = 2, name = "cbd1", alpha = 1)[[2]])
upvd
```



```{r}

########################################################################

regulation_da <- define_regulation(WT_vs_da)
UPregulation_da <- upregulation(regulation_da)
DOWNregulation_da <- downregulation(regulation_da)
regulation_2D <- define_regulation(WT_vs_2D)
UPregulation_2D <- upregulation(regulation_2D)
DOWNregulation_2D <- downregulation(regulation_2D)
regulation_sc <- define_regulation(WT_vs_sc)
UPregulation_sc <- upregulation(regulation_sc)
DOWNregulation_sc <- downregulation(regulation_sc)


complete_upregulation <- complete_regulation(UPregulation_da[-2],UPregulation_2D[-2],UPregulation_sc[-2])
complete_downregulation <- complete_regulation(DOWNregulation_da[-2],DOWNregulation_2D[-2],DOWNregulation_sc[-2])

# Triple upregulation
draw.triple.venn(area1 = nrow(subset(complete_upregulation, da == 1)), area2 = nrow(subset(complete_upregulation, DaDa == 1)), area3 = nrow(subset(complete_upregulation, sc == 1)), n12 = nrow(subset(complete_upregulation, da == 1 & DaDa == 1)), n23 = nrow(subset(complete_upregulation, DaDa == 1 & sc == 1)), n13 = nrow(subset(complete_upregulation,da == 1 & sc == 1)), n123 = nrow(subset(complete_upregulation, da == 1 & DaDa == 1 & sc ==1)), category = c("Da > 2 fold UP", "DaDa > 2 fold UP", "Sc > 2 fold UP"), lty = "blank", fill = c("skyblue", "pink1", "mediumorchid"))
# Triple downregulation
draw.triple.venn(area1 = nrow(subset(complete_downregulation, da == 1)), area2 = nrow(subset(complete_downregulation, DaDa == 1)), area3 = nrow(subset(complete_downregulation, sc == 1)), n12 = nrow(subset(complete_downregulation, da == 1 & DaDa == 1)), n23 = nrow(subset(complete_downregulation, DaDa == 1 & sc == 1)), n13 = nrow(subset(complete_downregulation,da == 1 & sc == 1)), n123 = nrow(subset(complete_downregulation, da == 1 & DaDa == 1 & sc ==1)), category = c("Da > 2 fold DOWN", "DaDa > 2 fold DOWN", "Sc > 2 fold DOWN"), lty = "blank", fill = c("skyblue", "pink1", "mediumorchid"))

# Sc upregulation DaDa downregulation
scUP_dadaDOWN <- merge(UPregulation_sc, DOWNregulation_2D, by=1, all=TRUE)
colnames(scUP_dadaDOWN) <- c("ensembleGeneID", "scUP","DaDaDOWN")
rownames(scUP_dadaDOWN) <- scUP_dadaDOWN$ensembleGeneID
scUP_dadaDOWN$ensembleGeneID <- NULL

draw.pairwise.venn(area1 = nrow(subset(scUP_dadaDOWN, scUP == 1)), area2 = nrow(subset(scUP_dadaDOWN, DaDaDOWN == 1)),cross.area = nrow(subset(scUP_dadaDOWN, scUP == 1 & DaDaDOWN == 1)), category = c("sc > 2 fold UP", "Da:Da > 2 fold DOWN"), lty = rep("blank",2), fill = c("light blue", "mediumorchid"), alpha = rep(0.5, 2), cat.pos = c(0,0), cat.dist = rep(0.025, 2), scaled = FALSE)
# Sc downregulation DaDa upregulation
scDOWN_dadaUP <- merge(DOWNregulation_sc, UPregulation_2D, by=1, all=TRUE)
colnames(scDOWN_dadaUP) <- c("ensembleGeneID", "scDOWN","DaDaUP")
rownames(scDOWN_dadaUP) <- scDOWN_dadaUP$ensembleGeneID
scDOWN_dadaUP$ensembleGeneID <- NULL

draw.pairwise.venn(area1 = nrow(subset(scDOWN_dadaUP, scDOWN == 1)), area2 = nrow(subset(scDOWN_dadaUP, DaDaUP == 1)),cross.area = nrow(subset(scDOWN_dadaUP, scDOWN == 1 & DaDaUP == 1)), category = c("sc > 2 fold DOWN", "Da:Da > 2 fold UP"), lty = rep("blank",2), fill = c("light blue", "mediumorchid"), alpha = rep(0.5, 2), cat.pos = c(0,0), cat.dist = rep(0.025, 2), scaled = FALSE)

# All conditions upregulation (default in the function)
plotConditions(c("d", "dd", "sc", "emc"), category = c("Da > 2 fold UP", "DaDa > 2 fold UP","sc > 2 fold UP", "emcRNAi > 2 fold UP"), lty = "blank", fill = c("skyblue", "pink1", "mediumorchid", "orange"))

# All conditions downregulation (change in the function)
plotConditions(c("d", "dd", "sc", "emc"), category = c("Da > 2 fold DOWN", "DaDa > 2 fold DOWN","sc > 2 fold DOWN", "emcRNAi > 2 fold DOWN"), lty = "blank", fill = c("skyblue", "pink1", "mediumorchid", "orange"), scaled = TRUE)



####################################################

complete_upregulation_ex <- add_rownames(complete_downregulation, "name")
dSets <- melt(complete_upregulation_ex[-5], id = "name")
dSets <- (subset(dSets, value == 1))[1:2]
plot(venneuler(dSets))


####################################################

upset(complete_upregulation[-5], sets = c("da", "DaDa", "sc"), sets.bar.color = "#56B4E9",
      order.by = "freq", empty.intersections = "on")
grid.text("> 2 fold upregulation",x = 0.65, y=0.95, gp=gpar(fontsize=18))


upset(complete_downregulation[-5], sets = c("da", "DaDa", "sc"), sets.bar.color = "#56B4E9",
      order.by = "freq", empty.intersections = "on")
grid.text("> 2 fold downregulation",x = 0.65, y=0.95, gp=gpar(fontsize=18))
```