---
title: "Visualisation of DESeq2 results with scatter plots"
description: "DEG analysis based on DESeq2 and GSEA"
principal investigator: "Joaquín de Navascués"
researcher: "Aleix Puig, modified by Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: 'hide'
    theme: readable
    df_print: paged
---
```{r set-publication-theme, echo=FALSE, cache=FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
```

```{r setup, echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 5, fig.width = 8.5, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path='integration/figures/', warning=FALSE, message=FALSE)
```

**Libraries needed:**
```{r load_libraries, warning=FALSE, echo=FALSE}
if (!require("librarian")) install.packages("librarian")
# data
librarian::shelf(readxl, dplyr, purrr, stringr, santoku, calibrate)
# graphics
librarian::shelf(RColorBrewer, cetcolor, ggplots2, ggpubr)
# convenience
librarian::shelf(here)
setwd(here::here()) # to distinguish from dplyr::here()
```

**Path to definitive images (outside repo):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(),'/')[[1]],-1),
                   paste0(tail(str_split(getwd(),'/')[[1]],1), '_figures')),
                 collapse='/')
dir.create(figdir, showWarnings = FALSE)
```


# RNAseq results visualisation with scatter plots


## 1 Getting ready


This gets us the DGE data from `DESeq2`, identified by FlyBase/Ensembl ID and gene symbol:
```{r load_DEG_data}
# experimental design and labels
targets <- readRDS('output/targets.RDS')
# DEG data
DaKD_deg <- readRDS('output/Control_vs_DaKD.RDS')
DaOE_deg <- readRDS('output/Control_vs_DaOE.RDS')
DaDaOE_deg <- readRDS('output/Control_vs_DaDaOE.RDS')
ScOE_deg <- readRDS('output/Control_vs_ScOE.RDS')
head(ScOE_deg, 1)
```

As discussed previously, we do not really need TPM values here, as the `baseMean` from the `DESeq2` normalisation does the work just as well and .
Calculate the `meanTpm` per condition:
# ```{r, calc_tpm}
# # get TPM means by condition, as a list of named vectors
# deg.conditions <- list("DaKD", "DaOE", "DaDaOE", "ScOE", "Control")
# tpmMeans <- lapply(deg.conditions, function(x)
#   rowMeans(tpm_bcorr[, targets[targets$Condition==x,'sampleIDs']]) )
# names(tpmMeans) <- deg.conditions
# head(tpmMeans[[1]])
# ```

Add it to the DEG data and remove unwanted columns:
```{r, data_cleanup}
# make ensembl IDs *also* rownames, remove unwanted cols
wanted_cols <- c('gene_symbol','log2FoldChange','padj', 'ensemblGeneID', 'baseMean')
cleanup <- function(df, wanted_cols, condition, control_tmp) {
  df <- df[, wanted_cols]
  df2 <- data.frame(tpmMeans[condition],
                    ensemblGeneID = names(tpmMeans[condition][[1]]) )
  names(df2)[[1]] <- 'baseMean'
  df3 <- data.frame(baseMean_ref  = tpmMeans$Control,
                    ensemblGeneID = names(tpmMeans[condition][[1]]) )
  
  df <- list(df, df2, df3) %>%
               reduce(full_join, by='ensemblGeneID')
  # df <- merge(df, df2, by=0)
  # df <- merge(df, , by=0)
  rownames(df) <- df$Row.names
  df$Row.names <- NULL
  return(df)
}
DaKD_deg   <- cleanup(DaKD_deg, wanted_cols, 'DaKD', tpmMeans$Control)
DaOE_deg   <- cleanup(DaOE_deg, wanted_cols, 'DaOE', tpmMeans$Control)
DaDaOE_deg <- cleanup(DaDaOE_deg, wanted_cols, 'DaDaOE', tpmMeans$Control)
ScOE_deg   <- cleanup(ScOE_deg, wanted_cols, 'ScOE', tpmMeans$Control)

head(DaKD_deg)
```

Now make a dataframe with the status of each gene (up/down/non-regulated) for each condition:
```{r extract_regulated_sets2, warning=FALSE}
# first define the function
extract_regulated_sets2 <- function(list_of_degs, names_degs, fc_thresh=2, cols) {
  # fc_thresh must be a positive number
  # list_of_degs is a list of dataframes
  # names_degs is a list of strings
  # they must have the same length
  reg_lvl <- paste0(' reg@log~2~FC≥', fc_thresh)
  breaks <- brk_manual(c(-fc_thresh, fc_thresh), left_vec = c(FALSE, TRUE))
  regulated_sets <- NULL
  for (l in 1:length(list_of_degs)) {
    degs_na <- dplyr::select(list_of_degs[[l]], all_of(cols))
    # better not pass NAs to kiru
    degs <- na.omit(degs_na)
    # filter by log2FC threshold, create `reg`(ulated) col with up/down/non
    degs$reg <- 
      kiru(
        degs$log2FoldChange,
        breaks = breaks,
        extend = TRUE,
        labels=c("down", "non", "up")
      )
    # filter by p-val
    degs$reg <- as.character(degs$reg) # remove factor
    x <- 1:nrow(degs)
    degs$reg <- ifelse(degs[x,'padj']<0.05, degs[x,'reg'], 'non')
    # recover NAs as non-regulataed
    added_nas <- degs_na[!(rownames(degs_na) %in% rownames(na.omit(degs_na))),]
    added_nas$reg <- 'non'
    degs <- rbind(degs, added_nas)
    # store reg status
    degs[ , names_degs[[l]] ] <- degs$reg
    regulated_sets[[l]] <- dplyr::select(degs, c(ensemblGeneID, names_degs[[l]]) )
  }
  regulated_sets <- purrr::reduce(regulated_sets, full_join, by='ensemblGeneID')
  rownames(regulated_sets) <- regulated_sets$ensemblGeneID
  regulated_sets <- dplyr::select(regulated_sets, -ensemblGeneID)
  return(regulated_sets)
}

# now use the function 
list_of_degs <- list('kd'=DaKD_deg, 'da'=DaOE_deg,
                     'dada'=DaDaOE_deg, 'sc'=ScOE_deg)
names_degs <- list('DaKD','DaOE','DaDaOE','ScOE')
cols.oi <- c('log2FoldChange', 'padj', 'meanTpm',
             'gene_symbol', 'ensemblGeneID')
regs <- extract_regulated_sets2(list_of_degs, names_degs, fc_thresh=2, cols=cols.oi)
head(regs)
```

Load `genes_of_interest`:
```{r load_GOI}
genes_of_interest <- read_excel('resources/genes_of_interest.xlsx')
```


## 2 Volcano plots


### First version


```{r plot_volcano_1, fig.height=5}
#Define table to plot
df <- list_of_degs$kd

# Make a basic volcano plot
with(subset(df, padj>=.05 | abs(log2FoldChange)<=1),
     plot(log2FoldChange, -log10(padj),
          pch=20, cex=.35, col="#999999",
          main=list("sc", cex=3.5),
          xlim=c( min(df$log2FoldChange, na.rm = TRUE),
                  max(df$log2FoldChange, na.rm = TRUE) ),
          ylim=c( 0,
                  max(-log10(subset(df, padj > 0)$padj), na.rm = TRUE) )
          )
     )

# Add colored points: red if padj<0.05, orange of log2FC>1, green if both)
with(subset(df, padj<.05 & abs(log2FoldChange)>1),
     points(log2FoldChange, -log10(padj),
            pch=20, cex=.35, col="#e2aa3b"))
with(filter(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange<(-1)),
     points(log2FoldChange, -log10(padj),
            pch=20, cex=.9, col="#721f0f"))
with(filter(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange>1),
     points(log2FoldChange, -log10(padj),
            pch=20, cex=.9, col="#18631a"))

# Label points with the textxy function from the calibrate plot
with(subset(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange<(-1)),
     textxy(log2FoldChange, -log10(padj),volv
            labs=external_gene_name,
            cex=0.9, font=2, col="#721f0f"))
with(subset(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange>1),
     textxy(log2FoldChange, -log10(padj),
            labs=external_gene_name,
            cex=0.9, font=2, col="#18631a"))
```


### Second version


```{r}

# Make a basic volcano plot
with(subset(df, padj>=.05 | abs(log2FoldChange)<=1),
     plot(log10(mean_tpm),
          log2FoldChange,
          pch=20, cex=.35, col=("#99999965"),
          main=list("Sc", cex=3.1),
          ylim=c(min(df$log2FoldChange, na.rm = TRUE),
                 max(df$log2FoldChange, na.rm = TRUE))))

# Add colored points: red if padj<0.05, orange of log2FC>1, green if both)
with(subset(df, padj<.05 & abs(log2FoldChange)>1),
     points(log10(mean_tpm),
            log2FoldChange,
            pch=20, cex=.35, col="#e2aa3b"))

with(filter(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange<(-1)),
     points(log10(mean_tpm),
            log2FoldChange,
            pch=20, cex=.9, col="#721f0f"))

with(filter(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange>1),
     points(log10(mean_tpm),
            log2FoldChange,
            pch=20, cex=.9, col="#18631a"))

abline(h=c(1,-1), lwd=c(2, 2))

# Label points with the textxy function from the calibrate plot

with(subset(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange<(-1)),
     textxy(log10(mean_tpm),
            log2FoldChange,
            labs=external_gene_name,
            cex=0.9, font=2, col="#721f0f"))
plot()

with(subset(df, external_gene_name %in% selected_genes & padj<.05 & log2FoldChange>1),
     textxy(log10(mean_tpm),
            log2FoldChange,
            labs=external_gene_name,
            cex=0.9, font=2, col="#18631a"))

```

### Third version, with the `EnhancedVolcano` package

https://www.bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html

### Fourth version: just with `ggplot2` and `ggrepel`

https://github.com/erikaduan/r_tips/blob/master/tutorials/dv-volcano_plots_with_ggplot/dv-volcano_plots_with_ggplot.md
https://rpkgs.datanovia.com/ggpubr/reference/ggmaplot.html

this could also work for the MA plot


## MA plots