---
title: "Analysis of DESeq2 results with gene set enrichment"
description: "DEG analysis based on DESeq2 and GSEA"
principal investigator: "Joaquín de Navascués"
researcher: "Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: 'hide'
    theme: readable
    df_print: paged
---

# Analysis of DESeq2 results with gene set enrichment

## 0 Boilerplate

```{r set-publication-theme, echo=FALSE, cache=FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
```

```{r setup, echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 7, fig.width = 8.5, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path='integration/figures/', warning=FALSE, message=FALSE)
```

**Libraries and external code needed:**
```{r load_libraries, warning=FALSE, echo=FALSE}
if (!require("librarian")) install.packages("librarian")
# data
librarian::shelf(
  # data manip.
  dplyr, tidyr, stringr, org.Dm.eg.db, DOSE, fgsea, clusterProfiler, purrr, biomaRt, gage,
  # plotting
  pathview,  cetcolor, enrichplot, genekitr, ggh4x, ggtheme, ggtext, ggrepel, ggnewscale, patchwork, pathview,
# convenience
  here)

if(!exists("gseCP_summarise", mode="function")) source("utils.R")
```

**Set working directory where the script is**
```{r setwd}
if (Sys.getenv("USER")=="JQ") {
  setwd("/Users/JQ/Documents/_CODE REPOS/GitHub/Da_RNAseq")
} else if (Sys.getenv("RSTUDIO")==1) {
  setwd( dirname(rstudioapi::getSourceEditorContext(id = NULL)$path) ) # gets what is in the editor
} else {
  setwd(here::here())
  d <- str_split(getwd(),'/')[[1]][length(str_split(getwd(),'/')[[1]])]
  if (d != 'Da_RNAseq') { stop(
    paste0("Could not set working directory automatically to where this",
           " script resides.\nPlease do `setwd()` manually"))
    }
}
getwd()
```

**Path to definitive images (outside repo):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(),'/')[[1]],-1),
                   paste0(tail(str_split(getwd(),'/')[[1]],1), '_figures')),
                 collapse='/')
dir.create(figdir, showWarnings = FALSE)
```


## 1 Getting ready


This gets us the DGE data from `DESeq2`, identified by FlyBase/Ensembl ID and gene symbol:
```{r load_DEG_data}
# experimental design and labels
targets <- readRDS('output/targets.RDS')
# DEG data
DaKD_deg <- readRDS('output/Control_vs_DaKD.RDS')
DaOE_deg <- readRDS('output/Control_vs_DaOE.RDS')
DaDaOE_deg <- readRDS('output/Control_vs_DaDaOE.RDS')
ScOE_deg <- readRDS('output/Control_vs_ScOE.RDS')
head(
  ScOE_deg %>% dplyr::select(gene_symbol, ensemblGeneID, baseMean, log2FoldChange, padj),
  1
)
```

#### Collect NCBI gene IDs to match KEGG database

This gets us the NCBI gene IDs (a.k.a. entrez gene ID):
```{r get_ncbi-geneid_keys}
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL",
                     dataset="dmelanogaster_gene_ensembl",
                     host = "https://oct2022.archive.ensembl.org")
attributes <- listAttributes(ensembl)
ezlist <- getBM(attributes=c('entrezgene_id', 'ensembl_gene_id', 'external_gene_name'), mart = ensembl)
ezlist <- drop_na(ezlist)
```

This gets us the KEGG sets for _Drosophila_:
```{r get_kegg_sets}
kegg.sets.dme <- kegg.gsets(species='dme', id.type='entrez')
```

Let us now separate signaling and metabolic pathways (easier to plot later)
```{r tidy_kegg_sets_sig}
# separate sig.naling vs met.abolic pathways and tidy
kegg.sig_tidy <- NULL
KSIG <- kegg.sets.dme$kg.sets[kegg.sets.dme$sig.idx]
for ( x in 1:length(KSIG) ) {
  df <- data.frame(gene = KSIG[[x]])
  df$term <- names(KSIG)[[x]]
  kegg.sig_tidy <- rbind(kegg.sig_tidy, dplyr::select(df, term, gene))
}
# split term names into KEGG ID and description
kegg.sig_tidy <- kegg.sig_tidy %>%
  separate_wider_delim(term, ' ', names = c('term', 'description'),
                       too_many = 'merge')
# show
kegg.sig_tidy[as.integer(seq(from=1, to=nrow(kegg.sig_tidy), length.out = 10)),]
```

Now metabolic ones:
Let us now separate signaling and metabolic pathways (easier to plot later)
```{r tidy_kegg_sets_met}
# separate sig.naling vs met.abolic pathways and tidy
kegg.met_tidy <- NULL
KMET <- kegg.sets.dme$kg.sets[kegg.sets.dme$met.idx]
for ( x in 1:length(KMET) ) {
  df <- data.frame(gene = KMET[[x]])
  df$term <- names(KMET)[[x]]
  kegg.met_tidy <- rbind(kegg.met_tidy, dplyr::select(df, term, gene))
}
# split term names into KEGG ID and description
kegg.met_tidy <- kegg.met_tidy %>%
  separate_wider_delim(term, ' ', names = c('term', 'description'),
                       too_many = 'merge')
kegg.met_tidy[as.integer(seq(from=1, to=nrow(kegg.met_tidy), length.out = 10)),]
```


## 2 Metabolic signatures using KEGG


To simplify the calls:
```{r}
# common parameters
kegg.sig_gmx <- kegg.sig_tidy %>% dplyr::select(description, gene)
names(kegg.sig_gmx) <- c('term', 'gene')
GSEAparams_sig <- list(exponent = 1, minGSSize = 1, maxGSSize = 5000, eps = 0, pvalueCutoff = 1,
                       pAdjustMethod = "BH", TERM2GENE = kegg.sig_gmx, TERM2NAME = NA, verbose = TRUE,
                       seed = FALSE, by = "fgsea")

kegg.met_gmx <- kegg.met_tidy %>% dplyr::select(description, gene)
names(kegg.met_gmx) <- c('term', 'gene')
GSEAparams_met <- list(exponent = 1, minGSSize = 1, maxGSSize = 5000, eps = 0, pvalueCutoff = 1,
                       pAdjustMethod = "BH", TERM2GENE = kegg.met_gmx, TERM2NAME = NA, verbose = TRUE,
                       seed = FALSE, by = "fgsea")
```

Now we can do, for each condition with DEG data, GSEA using the signaling or metabolic pathways as gene set collections:
```{r, message=FALSE}
# create rank with ncbi-geneid keys instead of gene symbols
daKD_rank <- make_degrank(DaKD_deg, mode='log2fc', key='gene_symbol')
daKD_kegg_rank <- daKD_rank 
names(daKD_kegg_rank) <- ezlist[match(names(daKD_rank), ezlist$external_gene_name), 'entrezgene_id']
# some names will be NA because they are not in the NCBI db (?), so to filter them:
daKD_kegg_rank <- daKD_kegg_rank[-na.action(na.omit(names(daKD_kegg_rank)))]
# now we can do the GSEA safely
daKD_gse_ksig   <- do.call(GSEA, c(list(geneList=daKD_kegg_rank),   GSEAparams_sig))
daKD_gse_kmet   <- do.call(GSEA, c(list(geneList=daKD_kegg_rank),   GSEAparams_met))

# prep
daOE_rank <- make_degrank(DaOE_deg, mode='log2fc', key='gene_symbol')
daOE_kegg_rank <- daOE_rank 
names(daOE_kegg_rank) <- ezlist[match(names(daOE_rank), ezlist$external_gene_name), 'entrezgene_id']
daOE_kegg_rank <- daOE_kegg_rank[-na.action(na.omit(names(daOE_kegg_rank)))]
# GSEA
daOE_gse_ksig   <- do.call(GSEA, c(list(geneList=daOE_kegg_rank),   GSEAparams_sig))
daOE_gse_kmet   <- do.call(GSEA, c(list(geneList=daOE_kegg_rank),   GSEAparams_met))

#prep
dadaOE_rank <- make_degrank(DaDaOE_deg, mode='log2fc', key='gene_symbol')
dadaOE_kegg_rank <- dadaOE_rank 
names(dadaOE_kegg_rank) <- ezlist[match(names(dadaOE_rank), ezlist$external_gene_name), 'entrezgene_id']
dadaOE_kegg_rank <- dadaOE_kegg_rank[-na.action(na.omit(names(dadaOE_kegg_rank)))]
# GSEA
dadaOE_gse_ksig <- do.call(GSEA, c(list(geneList=dadaOE_kegg_rank), GSEAparams_sig))
dadaOE_gse_kmet <- do.call(GSEA, c(list(geneList=dadaOE_kegg_rank), GSEAparams_met))

#prep
scOE_rank <- make_degrank(ScOE_deg, mode='log2fc', key='gene_symbol')
scOE_kegg_rank <- scOE_rank 
names(scOE_kegg_rank) <- ezlist[match(names(scOE_rank), ezlist$external_gene_name), 'entrezgene_id']
scOE_kegg_rank <- scOE_kegg_rank[-na.action(na.omit(names(scOE_kegg_rank)))]
# GSEA
scOE_gse_ksig   <- do.call(GSEA, c(list(geneList=scOE_kegg_rank),   GSEAparams_sig))
scOE_gse_kmet   <- do.call(GSEA, c(list(geneList=scOE_kegg_rank),   GSEAparams_met))
```

#### Layered heatmaps

First to produce the dataframes for `ggplot2`, for _signaling_ pathways:
```{r create_dfs_for_NESheatmap_sig}
gse_list <- list(scOE_gse_ksig, dadaOE_gse_ksig, daOE_gse_ksig, daKD_gse_ksig)
conditions <- c('*scute*', '*da:da*', '*da*', '*da^RNAi^*')
sets.as.factors <- unique(kegg.sig_tidy$description)
layerhm.df <- gseCP_summarise(kegg.sig_gmx, gse_list, conditions, sets.as.factors, cluster=TRUE, nsig.out = TRUE)

subt <- "for KEGG *signaling* pathways"
p <- layer.heatmap(layerhm.df, subt)
p + geom_hline(aes(yintercept=3.5), linewidth = 0.5) +
  theme(plot.margin = margin(r = 200))
```

Now _metabolic_ pathways:
```{r create_dfs_for_NESheatmap_met}
gse_list <- list(scOE_gse_kmet, dadaOE_gse_kmet, daOE_gse_kmet, daKD_gse_kmet)
# `conditions` were defined further above
sets.as.factors <- unique(kegg.met_tidy$description)
layerhm.df <- gseCP_summarise(kegg.met_gmx, gse_list, conditions, sets.as.factors, cluster=TRUE, nsig.out = TRUE)

subt <- "for KEGG *metabolic* pathways"
p <- layer.heatmap(layerhm.df, subt)
p + geom_hline(aes(yintercept=3.5), linewidth = 0.5) +
  theme(plot.margin = margin(r = 200))
```

#### Pathway plots

Obtain KEGG id for the desired pathways:
```{r get_keggids}
path.ids <- kegg.met_tidy %>%
  dplyr::select(term, description) %>%
  distinct()
```

To simplify calls
```{r handled_pathview_param}
# fixed parameters
fixed <- list(
  low = list(gene = cet_pal(3, name='cbd1')[1], cpd = "blue"),
  mid = list(gene = cet_pal(3, name='cbd1')[2], cpd = "gray"),
  high = list(gene = cet_pal(3, name='cbd1')[3], cpd = "yellow"),
  species = "dme",
  kegg.dir = figdir,
  new.signature=FALSE,
  res = 600)
```

Plot pathway for Tryptophan metabolism:
```{r example_pathview}
(path.ids %>% filter(description == 'Tryptophan metabolism'))$term
```
```{r}
params <- c(
  list(gene.data = cbind(daKD_rank, dadaOE_rank = dadaOE_rank[names(daKD_rank)]),
       pathway.id = "dme00380",
       out.suffix = "daRNAi_dada",
       gene.idtype = 'SYMBOL',
       limit = list(gene=1, cpd=1)
       ),
  fixed)
handled_pathview(params)
```
