---
title: "Analysis of DESeq2 results with gene set enrichment"
description: "DEG analysis based on DESeq2 and GSEA"
principal investigator: "Joaquín de Navascués"
researcher: "Joaquín de Navascués"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: 'hide'
    theme: readable
    df_print: paged
---
```{r set-publication-theme, echo=FALSE, cache=FALSE}
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
```

```{r setup, echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(dev = 'png', 
                      fig.align = 'center', fig.height = 7, fig.width = 8.5, 
                      pdf.options(encoding = "ISOLatin9.enc"),
                      fig.path='integration/figures/', warning=FALSE, message=FALSE)
```

**Libraries needed:**
```{r load_libraries, warning=FALSE, echo=FALSE}
if (!require("librarian")) install.packages("librarian")
# data
librarian::shelf(dplyr, org.Dm.eg.db, fgsea, DOSE, pathview, clusterProfiler)# purrr, stringr, santoku
# graphics
librarian::shelf(cetcolor, ggplot2, enrichplot, ggh4x, ggtheme, ggtext)#, ggrepel, ggpubr)
# convenience
librarian::shelf(here)
setwd(here::here()) # to distinguish from dplyr::here()
getwd()
```

**Path to definitive images (outside repo):**
```{r define_dir2figs}
figdir <- paste0(c(head(str_split(getwd(),'/')[[1]],-1),
                   paste0(tail(str_split(getwd(),'/')[[1]],1), '_figures')),
                 collapse='/')
dir.create(figdir, showWarnings = FALSE)
```


# Analysis of DESeq2 results with gene set enrichment


## 1 Getting ready


This gets us the DGE data from `DESeq2`, identified by FlyBase/Ensembl ID and gene symbol:
```{r load_DEG_data}
# experimental design and labels
targets <- readRDS('output/targets.RDS')
# DEG data
DaKD_deg <- readRDS('output/Control_vs_DaKD.RDS')
DaOE_deg <- readRDS('output/Control_vs_DaOE.RDS')
DaDaOE_deg <- readRDS('output/Control_vs_DaDaOE.RDS')
ScOE_deg <- readRDS('output/Control_vs_ScOE.RDS')
head(
  ScOE_deg %>% dplyr::select(gene_symbol, ensemblGeneID, baseMean, log2FoldChange, padj),
  1
)
```

To get a differentially expressed gene set for Over-Representation Analysis:
```{r define_degset}
degset_symbols <- function(deg, up, fc_thresh) {
  converter <- ifelse(up,1,-1)
  degset <- deg %>%
    filter(padj<0.05) %>%
    filter(abs(log2FoldChange) > fc_thresh) %>%
    filter(log2FoldChange*converter > 0)
 return(degset)
}
```

To get a ranked gene list for Gene Set Enrichment Analysis:
```{r define_degrank}
degrank_symbols <- function(deg, mode='log2fc') {
  # mode: whether to use padj or log2FC
  if (mode=='log2fc') {
    deg_sorted <- deg %>% arrange(desc(log2FoldChange))
    degrank <- deg_sorted$log2FoldChange
    names(degrank) <- deg_sorted$gene_symbol
  } else if (mode=='padj') {
    deg$p.rank <- sign(deg$log2FoldChange) * -log10(deg$padj)
    deg_sorted <- deg %>% arrange(desc(p.rank))
    degrank <- deg_sorted$p.rank
    names(degrank) <- deg_sorted$gene_symbol
  } else {
    stop('the `mode` parameter can only be either `log2fc` or `padj`')
  }
 return(degrank)
}
```


## 2 Over-representation of Gene Ontology terms 


For *daughterless* knockdown:
```{r ORA_da_knockdown}
# using `clusterProfiler` and `enrichplot`
# select reg set
daKD_up <- degset_select_symbols(DaKD_deg, up=TRUE, fc_thresh=1.5)
# ORA
daKD_up_eGO <- enrichGO(gene         = daKD_up$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
# with down genes
daKD_dn <- degset_select_symbols(DaKD_deg, up=FALSE, fc_thresh=1.5)
daKD_dn_eGO <- enrichGO(gene         = daKD_dn$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
```
```{r ORA_da_knockdown_plots}
# plot
up <- dotplot(daKD_up_eGO, showCategory=30, label_format=50) +
  ggtitle("Over-represented GO terms — genes **up** in *esg > da^RNAi^*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[3]) +
  theme(plot.title = element_markdown(size=rel(1.2)),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=8, lineheight=0.5))

down <- dotplot(daKD_dn_eGO, showCategory=30, label_format=40) +
  ggtitle("Over-represented GO terms — genes **down** in *esg > da^RNAi^*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[1]) +
  theme(plot.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9))

suppressWarnings(print(up))
suppressWarnings(print(down))
```

```{r dotplots_using_`genekitr`, include=FALSE}
# plotEnrich(importCP(daKD_up_eGO, type='go'),
#            plot_type = 'dot',
#            up_color = cet_pal(3,name='cbd1')[3],
#            down_color = cet_pal(3,name='cbd1')[2])
# 
# plotEnrich(importCP(daKD_dn_eGO, type='go'),
#            plot_type = 'dot',
#            up_color = cet_pal(3,name='cbd1')[1],
#            down_color = cet_pal(3,name='cbd1')[2])
# plotEnrichAdv(importCP(daKD_up_eGO, type='go'), importCP(daKD_dn_eGO, type='go'),
#               plot_type = 'one',
#               term_metric = "FoldEnrich",
#               stats_metric = "p.adjust",
#               color =c(cet_pal(2,name='cbd1')[1], cet_pal(2,name='cbd1')[2]))
```

For *daughterless* overexpression:
```{r ORA_da_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
daOE_up <- degset_select_symbols(DaOE_deg, up=TRUE, fc_thresh=1.5)
# ORA
daOE_up_eGO <- enrichGO(gene         = daOE_up$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
# with down genes
daOE_dn <- degset_select_symbols(DaOE_deg, up=FALSE, fc_thresh=1.5)
daOE_dn_eGO <- enrichGO(gene         = daOE_dn$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
```
```{r ORA_da_overexpression_plots}
# plot
up <- dotplot(daOE_up_eGO, showCategory=30, label_format=60) +
  ggtitle("Over-represented GO terms — genes **up** in *esg > da*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[3]) +
  theme(plot.title = element_markdown(size=rel(1.15)),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9))

down <- dotplot(daOE_dn_eGO, showCategory=30, label_format=60) +
  ggtitle("Over-represented GO terms — genes **down** in *esg > da*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[1]) +
  theme(plot.title = element_markdown(size=rel(1.15)),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=8, lineheight=0.5))

suppressWarnings(print(up))
suppressWarnings(print(down))
```

For *da:da* overexpression:
```{r ORA_dada_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
dadaOE_up <- degset_select_symbols(DaDaOE_deg, up=TRUE, fc_thresh=1.5)
# ORA
dadaOE_up_eGO <- enrichGO(gene         = dadaOE_up$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
# with down genes
dadaOE_dn <- degset_select_symbols(DaDaOE_deg, up=FALSE, fc_thresh=1.5)
dadaOE_dn_eGO <- enrichGO(gene         = dadaOE_dn$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
```
```{r ORA_dada_overexpression_plots}
# plot
up <- dotplot(dadaOE_up_eGO, showCategory=30, label_format=40) +
  ggtitle("Over-represented GO terms — genes **up** in *esg > da:da*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[3]) +
  force_panelsizes(rows = unit(3, "in"), cols = unit(2, "in")) +
  theme(plot.title = element_markdown(),
        plot.title.position = 'plot',
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9))

suppressWarnings(print(up))
```

For *scute* overexpression:
```{r ORA_sc_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
scOE_up <- degset_select_symbols(ScOE_deg, up=TRUE, fc_thresh=1.5)
# ORA
scOE_up_eGO <- enrichGO(gene         = scOE_up$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
# with down genes
scOE_dn <- degset_select_symbols(ScOE_deg, up=FALSE, fc_thresh=1.5)
scOE_dn_eGO <- enrichGO(gene         = scOE_dn$gene_symbol,
                        OrgDb         = org.Dm.eg.db,
                        keyType       = 'SYMBOL',
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.01,
                        qvalueCutoff  = 0.05)
```
```{r ORA_sc_overexpression_plots}
# plot
up <- dotplot(scOE_up_eGO, showCategory=30, label_format=60) +
  ggtitle("Over-represented GO terms — genes **up** in *esg > scute*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[3]) +
  theme(plot.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9))

down <- dotplot(scOE_dn_eGO, showCategory=30, label_format=60) +
  ggtitle("Over-represented GO terms — genes **down** in *esg > scute*") +
  scale_colour_gradient(low = cet_pal(3, name='cbd1')[2],
                        high = cet_pal(3, name='cbd1')[1]) +
  theme(plot.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=8, lineheight=0.5))

suppressWarnings(print(up))
suppressWarnings(print(down))
```

None of these plots are particularly informative, so I am going to move on to GSEA and custom gene sets.


## 3 Gene Set Enrichment Analysis with Gene Ontology terms


For *daughterless* knockdown:
```{r GSEA_da_knockdown}
# using `clusterProfiler` and `enrichplot`
# select reg set
daKD_rank <- degrank_symbols(DaKD_deg, mode='log2fc')
# GSEA
daKD_GSEgo <- gseGO(geneList=daKD_rank,
                    keyType = 'SYMBOL',
                    OrgDb=org.Dm.eg.db,
                    minGSSize    = 15,
                    maxGSSize    = 1000,
                    pvalueCutoff = 0.05,
                    eps = 0,
                    verbose=F)
```

```{r GSEA_da_knockdown_plots}
# get the logP values instead of padj 
daKD_GSEgo_log <- daKD_GSEgo
daKD_GSEgo_log@result$p.adjust <- -log10(daKD_GSEgo_log@result$p.adjust) # use @slot and $column to assign
# to subset the data to the highest NES values
#daKD_GSEgo_log@result <- subset(daKD_GSEgo_log@result, abs(NES)>1.8)
# to subset the data to the lowest p-values:
number_of_terms <- 30
cutoff <- rev(rev(sort(daKD_GSEgo_log$p.adjust))[1:number_of_terms])[1]
daKD_GSEgo_log@result <- subset(daKD_GSEgo_log@result, p.adjust>=cutoff)
# to control colour of GO terms according to enrichment or depletion
d <- ifelse(sort(daKD_GSEgo_log$NES) > 0, cet_pal(2, name='cbd1')[2], cet_pal(2, name='cbd1')[1])
# plot
gsep <- dotplot(daKD_GSEgo_log,
                x='NES', 
                showCategory=number_of_terms,
                label_format=60,
                ) +
  ggtitle("Enriched/depleted GO terms in *esg > da^RNAi^*") +
  scale_colour_gradient(low = cet_pal(3, name='d2')[2],
                        high = cet_pal(3, name='d2')[3]) +
  labs(color = '-log~10~(*p*)') +
  theme(plot.title = element_markdown(),
        plot.title.position = 'panel',
        legend.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9, colour=d))

suppressWarnings(print(gsep))
```

For *daughterless* overexpression:
```{r GSEA_da_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
daOE_rank <- degrank_symbols(DaOE_deg, mode='log2fc')
# GSEA
daOE_GSEgo <- gseGO(geneList=daOE_rank,
                    keyType = 'SYMBOL',
                    OrgDb=org.Dm.eg.db,
                    minGSSize    = 15,
                    maxGSSize    = 1000,
                    pvalueCutoff = 0.05,
                    eps = 0,
                    verbose=F)
```

```{r GSEA_da_overexpression_plots}
# get the logP values instead of padj 
daOE_GSEgo_log <- daOE_GSEgo
daOE_GSEgo_log@result$p.adjust <- -log10(daOE_GSEgo_log@result$p.adjust) # use @slot and $column to assign
# to subset the data to the highest NES values
#daOE_GSEgo_log@result <- subset(daOE_GSEgo_log@result, abs(NES)>1.8)
# to subset the data to the lowest p-values:
number_of_terms <- 30
cutoff <- rev(rev(sort(daOE_GSEgo_log$p.adjust))[1:number_of_terms])[1]
daOE_GSEgo_log@result <- subset(daOE_GSEgo_log@result, p.adjust>=cutoff)
# to control colour of GO terms according to enrichment or depletion
d <- ifelse(sort(daOE_GSEgo_log$NES) > 0, cet_pal(2, name='cbd1')[2], cet_pal(2, name='cbd1')[1])
# plot
gsep <- dotplot(daOE_GSEgo_log,
                x='NES', 
                showCategory=number_of_terms,
                label_format=60,
                ) +
  ggtitle("Enriched/depleted GO terms in *esg > da*") +
  scale_colour_gradient(low = cet_pal(3, name='d2')[2],
                        high = cet_pal(3, name='d2')[3]) +
  labs(color = '-log~10~(*p*)') +
  theme(plot.title = element_markdown(),
        plot.title.position = 'panel',
        legend.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9, colour=d))

suppressWarnings(print(gsep))
```

For *da:da* overexpression:
```{r GSEA_dada_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
dadaOE_rank <- degrank_symbols(DaDaOE_deg, mode='log2fc')
# GSEA
dadaOE_GSEgo <- gseGO(geneList=dadaOE_rank,
                    keyType = 'SYMBOL',
                    OrgDb=org.Dm.eg.db,
                    minGSSize    = 15,
                    maxGSSize    = 1000,
                    pvalueCutoff = 0.05,
                    eps = 0,
                    verbose=F)
```

```{r GSEA_da_overexpression_plots}
# get the logP values instead of padj 
dadaOE_GSEgo_log <- dadaOE_GSEgo
dadaOE_GSEgo_log@result$p.adjust <- -log10(dadaOE_GSEgo_log@result$p.adjust) # use @slot and $column to assign
# to subset the data to the highest NES values
#dadaOE_GSEgo_log@result <- subset(dadaOE_GSEgo_log@result, abs(NES)>1.8)
# to subset the data to the lowest p-values:
number_of_terms <- 30
cutoff <- rev(rev(sort(dadaOE_GSEgo_log$p.adjust))[1:number_of_terms])[1]
dadaOE_GSEgo_log@result <- subset(dadaOE_GSEgo_log@result, p.adjust>=cutoff)
# to control colour of GO terms according to enrichment or depletion
d <- ifelse(sort(dadaOE_GSEgo_log$NES) > 0,
            cet_pal(2, name='cbd1')[2], cet_pal(2, name='cbd1')[1])
# plot
gsep <- dotplot(dadaOE_GSEgo_log,
                x='NES', 
                showCategory=number_of_terms,
                label_format=50,
                ) +
  ggtitle("Enriched/depleted GO terms in *esg > da:da*") +
  scale_colour_gradient(low = cet_pal(3, name='d2')[2],
                        high = cet_pal(3, name='d2')[3]) +
  labs(color = '-log~10~(*p*)') +
  #coord_fixed(ratio=0.8) +
  force_panelsizes(rows = unit(3.5,'in'), cols = unit(4,'in')) +
  theme(plot.title = element_markdown(),
        plot.title.position = 'panel',
        legend.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=7, lineheight=0.7, colour=d)
        )
#gsep$coordinates$ratio <- 0.01
suppressWarnings(print(gsep))
```

For *scute* overexpression:
```{r GSEA_dada_overexpression}
# using `clusterProfiler` and `enrichplot`
# select reg set
scOE_rank <- degrank_symbols(ScOE_deg, mode='log2fc')
# GSEA
scOE_GSEgo <- gseGO(geneList=scOE_rank,
                    keyType = 'SYMBOL',
                    OrgDb=org.Dm.eg.db,
                    minGSSize    = 15,
                    maxGSSize    = 1000,
                    pvalueCutoff = 0.05,
                    nPermSimple = 10000,
                    eps = 0,
                    verbose=F)
```

```{r GSEA_da_overexpression_plots}
# get the logP values instead of padj 
scOE_GSEgo_log <- scOE_GSEgo
scOE_GSEgo_log@result$p.adjust <- -log10(scOE_GSEgo_log@result$p.adjust) # use @slot and $column to assign
# to subset the data to the highest NES values
#scOE_GSEgo_log@result <- subset(scOE_GSEgo_log@result, abs(NES)>1.8)
# to subset the data to the lowest p-values:
number_of_terms <- 30
cutoff <- rev(rev(sort(scOE_GSEgo_log$p.adjust))[1:number_of_terms])[1]
scOE_GSEgo_log@result <- subset(scOE_GSEgo_log@result, p.adjust>=cutoff)
# to control colour of GO terms according to enrichment or depletion
d <- ifelse(sort(scOE_GSEgo_log$NES) > 0,
            cet_pal(2, name='cbd1')[2], cet_pal(2, name='cbd1')[1])
# plot
gsep <- dotplot(scOE_GSEgo_log,
                x='NES', 
                showCategory=number_of_terms,
                label_format=60
                ) +
  ggtitle("Enriched/depleted GO terms in *esg > scute*") +
  scale_colour_gradient(low = cet_pal(3, name='d2')[2],
                        high = cet_pal(3, name='d2')[3]) +
  labs(color = '-log~10~(*p*)') +
  theme(plot.title = element_markdown(),
        plot.title.position = 'panel',
        legend.title = element_markdown(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size=9, colour=d))

suppressWarnings(print(gsep))
```


## 4 Enrichment analysis of custom gene sets


### 4.1 Cell-type specific transcriptomes (Dutta et al, Doupée et al)


### 4.2 RNAi screen phenotypes (Zhang et al)


### 4.3 Gene List Annotation for Drosophila (Mohr et al)


